\pkg{casebase}: An Alternative Framework for Survival Analysis and Comparison of Event Rates

### 2022-11-11 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
switch to distill

% !TeX root = RJwrapper.tex
\title{\pkg{casebase}: An Alternative Framework For Survival Analysis
and Comparison of Event Rates}
\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A.
Hanley, and Olli Saarela}

@@

\abstract{%
In clinical studies of time-to-event data, a quantity of interest to the
clinician is their patient's risk of an event. However, methods relying
on time matching or risk-set sampling (including Cox regression)
eliminate the baseline hazard from the estimating function. As a
consequence, the focus has been on reporting hazard ratios instead of
survival or cumulative incidence curves. Indeed, reporting patient risk
or cumulative incidence requires a separate estimation of the baseline
hazard. Using case-base sampling, Hanley \& Miettinen (2009) explained
how parametric hazard functions can be estimated in continuous-time
using logistic regression. Their approach naturally leads to estimates
of the survival or risk function that are smooth-in-time.\\
In this paper, we present the \CRANpkg{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can also be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis. We
illustrate all these features through three different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

The semiparametric Cox model has become the default approach to survival
analysis even though Cox himself later suggested he would prefer to
model the hazard function directly. In a 1994 interview with Professor
Nancy Reid, Sir David Cox was asked how he would model a set of censored
survival data, to which he responded: ``I think I would normally want to
tackle problems parametrically \ldots{} and if you want to do things
like predict the outcome for a particular patient, it's much more
convenient to do that parametrically'' \citep{reid1994conversation}.
Indeed, the most relevant quantity in a clinical setting is often the 5-
or 10-year risk of experiencing a certain event given the patient's
particular profile. However, the most reported metric from a Cox model
is the (potentially time-dependent) hazard ratio (HR). The
covariate-conditional survival curve is arguably a more important
summary measure to report than the HR. While stepwise survival curves
can be computed with the Cox model, they require a second step to
separately estimate the baseline hazard \citep{breslow1972discussion}.

Several authors have since pursued fully parametric approaches that made
the fitting of smooth survival curves more transparent and intuitive
through generalized linear models. The key feature of these procedures
is splitting the time axis into discrete intervals. Whitehead
\citeyearpar{whitehead1980fitting} showed the equivalence between a Cox
model and a Poisson regression with a parameter for each event time;
Carstensen \citeyearpar{copenhagen2012needs} provides an exposition of
this equivalence with a real data example and supporting R code for
computing standard errors. Arjas \& Haara
\citeyearpar{arjas1987logistic} and Efron
\citeyearpar{efron1988logistic} treated each patient-day as a Bernoulli
random variable with probability equal to the discrete hazard rate. A
potential issue with these approaches is that the number of time bins
need to be chosen by the data analyst. On the one hand, a fine grouping
of times may result in few (or none) events per interval, which then
leads to instability in the Newton-Raphson procedure for estimation
\citep[Section 4.8]{kalbfleisch2011statistical}, and large long-format
datasets. On the other hand, a coarse grouping could potentially mask
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} selected a discrete set of person-time
coordinates (``person-moments'') in continuous time from all observed
follow-up experience constituting the study base. By doing so, they
obtained a likelihood expression for the hazard function that is
equivalent to that of logistic regression with an offset term. More
specifically, all person-moments when the event of interest occurred are
selected as the case series, complemented by a randomly sampled base
series of person-moments serving as controls. This approach allows
flexible modeling of the hazard function by including functions of time
as covariates (e.g.~using splines or generalized additive models).
Furthermore, time-dependent covariates can be modeled through
interactions with time. In short, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} use the well-understood logistic
regression for directly modeling the hazard function, without requiring
a discrete-time model.

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} which implements and extends the Hanley \&
Miettinen \citeyearpar{hanley2009fitting} approach for fitting fully
parametric hazard models and covariate-conditional survival curves using
the familiar interface of the \code{glm} function. Our implementation
includes extensions to other models such as penalized regression for
variable selection and competing risk analysis. In addition, we provide
functions for exploratory data analysis and visualizing the estimated
quantities such as the hazard function, survival curve, and their
standard errors. The ultimate goal of our package is to make fitting
flexible hazards accessible to end users who favor reporting absolute
risks and survival curves over hazard ratios.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to consider the entire study base
as an infinite collection of \emph{person-moments}. These person-moments
are indexed by both an individual in the study and a time point, and
therefore each person-moment has a covariate profile and an outcome
status (i.e.~whether the event happened) attached to it. By estimating
the probability of an event occurring at a particular person-moment, we
can extract information about hazards and survival.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
independent censoring is treated in Saarela
\citeyearpar{saarela2016case}). We are interested in modeling the hazard
functions \(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which
satisfy
\[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\]
The processes \(N_i(t)\) count the person-moments from the \emph{case
series}.

@@

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
terms of \(\theta\), we can define an estimator \(\hat{\theta}\) by
maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta)\,\textrm dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{\,\textrm dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
\(u\), and where \(t_i\) is the event time for individual \(i\).
However, the integral over time makes the computation and maximization
of \(L_0(\theta)\) challenging using standard software.

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual partial likelihood
contributions of the form
\[P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
By using the symbol \(\stackrel{\theta}{\propto}\), we mean that both
sides of the expression are equal up to multiplicative factors that do
not depend on \(\theta\). Therefore, an estimating function for
\(\theta\) can be composed of these contributions as: \begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation} When a logarithmic link function is used for modeling the
hazard function, the above expression is of a logistic regression form
with an offset term \(\log(1/\rho_i(t))\). Note that the sampling units
selected in the case-base sampling mechanism are person-moments, rather
than individuals, and the parameters to be estimated are hazards or
hazard ratios rather than odds or odds ratios. Generally, an individual
can contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the corresponding partial
likelihood score function has mean zero at the true value
\(\theta=\theta_0\), and that the resulting estimator \(\hat{\theta}\)
is asymptotically normally distributed.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, their sampling strategy corresponds to a
time-homogeneous Poisson process with intensity equal to
\(\rho_i(t) = b/B\), where \(b\) is the number of sampled base series
\emph{person-moments}, and \(B\) is the total population-time for the
study base (e.g.~the sum of all individual follow-up times). More
complex examples are also possible; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the intensity functions for the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. Non-uniform sampling mechanisms can
increase the efficiency of the resulting maximum partial likelihood
estimators.

@@

The asymptotic normality of \(\hat{\theta}\) gives us an efficient way
to estimate the variance of various estimators derived from the hazard
function. For example, to construct confidence bands around risk
functions and survival functions, we can use the following procedure:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute \(\hat{\theta}\) and its variance-covariance matrix
  \(V(\hat \theta)\) which is obtained as part of the logistic
  regression output.
\item
  Sample \(B\) times from a multivariate normal
  \(N\left(\hat{\theta}, V(\hat\theta)\right)\).
\item
  For each sample, compute the survival function using Equation
  \ref{eqn:surv} below.
\item
  Use the pointwise quantiles of these survival function estimates to
  construct a pointwise confidence band for the survival function of
  interest.
\end{enumerate}

This procedure is similar to parametric bootstrap \citep[Section
6.5]{efron1994introduction}, but it can be more accurately described as
a form of approximate Bayesian computation. As such, the validity of the
confidence bands relies on the Bernstein-von Mises theorem
\citep[Chapter 10]{van2000asymptotic}.

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation} In this model, the instantaneous failure rate is
independent of
\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Use of \(\log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time \citep{kalbfleisch2011statistical}:

@@

\[\frac{\lambda_j(t)^{\,\textrm dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific
hazard function. As may be expected, this functional form is similar to
the terms appearing in the likelihood function for multinomial
regression with an offset
term.\footnote{Specifically, it corresponds to the following parametrization  for a multinomial random variable $Y$: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}

@@

Survival analysis is an important branch of applied statistics and
epidemiology. Accordingly, there is already a vast ecosystem of R
packages implementing different methodologies. In this section, we
describe how the functionalities of \pkg{casebase} compare to these
packages.

At the time of writing, a cursory examination of CRAN's \ctv{Survival}
Task View reveals that there are over 250 packages related to survival
analysis \citep{survTaskView}. For the purposes of this article, we
restricted our review to packages that implement at least one of the
following features: parametric modeling, non-proportional hazard models,
competing risk analysis, penalized estimation, and Cumulative Incidence
(CI) estimation. By searching for appropriate keywords in the
\code{DESCRIPTION} file of these packages, we found 60 relevant
packages. These 60 packages were then manually examined to determine
which ones are comparable to \pkg{casebase}. In particular, we excluded
packages that were focused on a different set of problems, such as
frailty and multistate models. The remaining 14 packages appear in Table
\ref{tab:surv-pkgs}, along with some of the functionalities they offer.

Parametric survival models are implemented in several packages, each
differing in the parametric distributions available: \CRANpkg{CFC}
\citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
\citeyearpar{flexsurv}, \CRANpkg{SmoothHazard}
\citeyearpar{smoothHazard}, \CRANpkg{rstpm2} \citeyearpar{clements_liu},
\CRANpkg{mets} \citeyearpar{scheike2014estimating}, and
\CRANpkg{survival} \citeyearpar{survival-package}. For example,
\pkg{SmoothHazard} is limited to Weibull distributions
\citeyearpar{smoothHazard}, whereas both \pkg{flexsurv} and
\pkg{survival} allow users to supply any distribution of their choice.
\pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can
model the effect of time using splines, which allows flexible modeling
of the hazard function. As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear
combination of covariates (including time). Therefore, our package is
more general in that it allows the user to model any linear or
non-linear transformation of time including splines and higher order
polynomials. Also, by including interaction terms between covariates and
time, it also allows users to fit (non-proportional) time-varying
coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
model any shape parameter.

Several packages implement penalized estimation for the Cox model:
\CRANpkg{glmnet} \citeyearpar{regpathcox}, \CRANpkg{glmpath}
\citeyearpar{park_hastie}, \CRANpkg{penalized} \citeyearpar{l1penal},
\CRANpkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{rstpm2} \citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation of the hazard function. To our knowledge,
\pkg{casebase} and \pkg{rstpm2} are the only packages to offer this
functionality.

Next, several R packages implement methodologies for competing risk
analysis; for a different perspective on this topic, see Mahani \&
Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
provides functionality for competing risk analysis and multistate
modelling. The package \CRANpkg{cmprsk} provides methods for
cause-specific subdistribution hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIs.

@@

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{ccccccccc}
\toprule
& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Risk}  \\
\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Estimates}           \\
\bottomrule
\textbf{casebase}        & \checkmark                        & \checkmark                         & \checkmark                     & \checkmark                & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
\textbf{CFC}             & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
\textbf{cmprsk}             & \checkmark                        &                           &                       &                  &                     &  \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{crrp}            & \checkmark                        &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{fastcox}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{flexrsurv}       &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
\textbf{flexsurv}        & \checkmark                        & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &  \checkmark              \\ \hline
\textbf{glmpath}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{mets}            & \checkmark                        &                           &                       & \checkmark                &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{penalized}       &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{riskRegression}  & \checkmark                         &                           & \checkmark                     &                  &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{rstpm2}          &                          & \checkmark                         &                      & \checkmark                & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                         \\ \hline
\textbf{SmoothHazard}    &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          & \checkmark                            &                                      \\ \hline
\textbf{survival}        & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                               \\ 
\bottomrule
\end{tabular}%
}
\caption{Comparison of various R packages for survival analysis. \textbf{Competing Risks}: whether an implementation for competing risks is present. \textbf{Allows Non PH}: includes models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: allows a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Risk estimates}: estimation of survival curve and cumulative incidence is available.}
\label{tab:surv-pkgs}
\end{table}

@@

The functions in the \pkg{casebase} package can be divided into two
categories: 1) exploratory data analysis, in the form of population-time
plots; and 2) parametric modeling of the hazard function. We strove for
compatibility with both \code{data.frame}s and \code{data.table}s; this
can be seen in the coding choices we made and the unit tests we wrote.

@@

By separating the sampling and estimation functions, we allow the
possibility of users implementing more complex sampling scheme (as
described in Saarela \citeyearpar{saarela2016case}), or more complex
study designs (e.g.~time-varying exposure).

>>>>

% !TeX root = RJwrapper.tex
\title{\pkg{casebase}: An Alternative Framework for Survival Analysis and Comparison of Event Rates}
\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A. Hanley, and Olli Saarela}

@@

\abstract{%
In clinical studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function. As a consequence, the focus has been on reporting hazard ratios instead of survival or cumulative incidence curves. Indeed, reporting patient risk or cumulative incidence requires a separate estimation of the baseline hazard. Using case-base sampling, Hanley \& Miettinen (2009) explained how parametric hazard functions can be estimated in continuous-time using logistic regression. Their approach naturally leads to estimates of the survival or risk function that are smooth-in-time. In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can also be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies. * SRB and MT contributed equally to this work.
}

@@

The semiparametric Cox model has become the default approach to survival analysis
even though Cox himself later suggested he would prefer to model the hazard function
directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked
how he would model a set of censored survival data, to which he responded: ``I
think I would normally want to tackle problems parametrically \ldots{} and if you want
to do things like predict the outcome for a particular patient, it's much more
convenient to do that parametrically'' (Reid 1994). Indeed, the most
relevant quantity in a clinical setting is often the 5- or 10-year risk of
experiencing a certain event given the patient's particular profile.
However, the most reported metric from a Cox model is the (potentially
time-dependent) hazard ratio (HR). The covariate-conditional
survival curve is arguably a more important
summary measure to report than the HR. While stepwise survival curves can be
computed with the Cox model, they require a second step to separately estimate
the baseline hazard (Breslow 1972).

Several authors have since pursued fully parametric approaches that made the
fitting of smooth survival curves more transparent and intuitive through
generalized linear models. The key feature of these procedures is splitting the
time axis into discrete intervals. Whitehead (1980) showed the
equivalence between a Cox model and a Poisson regression with a parameter for
each event time; Carstensen (2019) provides an exposition of this
equivalence with a real data example and supporting R code for computing
standard errors. Arjas \& Haara (1987) and Efron (1988) treated each
patient-day as a Bernoulli random variable with probability equal to the
discrete hazard rate. A potential issue with these approaches is that the
number of time bins need to be chosen by the data analyst. On the one hand, a fine grouping
of times may result in few (or none) events per interval, which then leads
to instability in the Newton-Raphson procedure for estimation
(Kalbfleisch and Prentice 2011, vol. 360, sec. 4.8), and large long-format datasets.
On the other hand, a coarse grouping could potentially mask
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley \& Miettinen (2009)
selected a discrete set of person-time coordinates (``person-moments'') in
continuous time from all observed follow-up experience constituting the
study base. By doing so, they obtained a likelihood expression for the hazard
function that is equivalent to that of logistic regression with an offset term.
More specifically, all person-moments when the event of interest
occurred are selected as the case series, complemented by a randomly sampled
base series of person-moments serving as controls. This approach allows
flexible modeling of the hazard function by including functions of time as covariates
(e.g.~using splines or generalized additive models). Furthermore, time-dependent
covariates can be modeled through interactions with time. In short, Hanley \&
Miettinen (2009) use the well-understood logistic regression
for directly modeling the hazard function, without requiring a discrete-time model.

In this article, we present the \CRANpkg{casebase} R package (Bhatnagar et al. 2021)
which implements and extends the Hanley \& Miettinen (2009) approach for
fitting fully parametric hazard models and covariate-conditional survival curves
using the familiar interface of the \texttt{glm} function. Our implementation
includes extensions to other models such as penalized
regression for variable selection and competing risk analysis. In addition,
we provide functions for exploratory data analysis and visualizing the estimated
quantities such as the hazard function, survival curve, and their standard errors.
The ultimate goal of our package is to make fitting flexible hazards accessible
to end users who favor reporting absolute risks and survival curves
over hazard ratios.

@@

As discussed in Hanley \& Miettinen (2009), the key idea behind case-base sampling is to consider the entire study base as an infinite collection of \emph{person-moments}. These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile and an outcome status (i.e.~whether the event happened) attached to it. By estimating the probability of an event occurring at a particular person-moment, we can extract information about hazards and survival.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of \(\sigma\)-algebras. Instead, following Aalen \textit{et al} (2008), we use the placeholder ``past'' to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela \& Arjas (2015) and Saarela (2016). First, let \(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the event of interest for individual \(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due to the end of follow-up at time \(\tau\) (the general case of independent censoring is treated in Saarela (2016)). We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\] The processes \(N_i(t)\) count the person-moments from the \emph{case series}.

@@

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in terms of \(\theta\), we can define an estimator \(\widehat{\theta}\) by maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta)\,\textrm dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{\,\textrm dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to \(u\), and where \(t_i\) is the event time for individual \(i\). However, the integral over time makes the computation and maximization of \(L_0(\theta)\) challenging using standard software.

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual partial likelihood contributions of the form
\[P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
By using the symbol \(\stackrel{\theta}{\propto}\), we mean that both sides of the expression are equal up to multiplicative factors that do not depend on \(\theta\). Therefore, an estimating function for \(\theta\) can be composed of these contributions as:
\begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation}
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela (2016) showed that the corresponding partial likelihood score function has mean zero at the true value \(\theta=\theta_0\), and that the resulting estimator \(\widehat{\theta}\) is asymptotically normally distributed.

In Hanley \& Miettinen (2009), the authors suggest sampling the base series \emph{uniformly} from the study base. In terms of Poisson processes, their sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to \(\rho_i(t) = b/B\), where \(b\) is the number of sampled base series \emph{person-moments}, and \(B\) is the total population-time for the study base (e.g.~the sum of all individual follow-up times). More complex examples are also possible; see for example Saarela \& Arjas (2015), where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score. Non-uniform sampling mechanisms can increase the efficiency of the resulting maximum partial likelihood estimators.

@@

The asymptotic normality of \(\widehat{\theta}\) gives us an efficient way to estimate the variance of various estimators derived from the hazard function. For example, to construct confidence bands around risk functions and survival functions, we can use the following procedure:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute \(\widehat{\theta}\) and its variance-covariance matrix \(V(\widehat \theta)\) which is obtained as part of the logistic regression output.
\item
  Sample \(B\) times from a multivariate normal \(N\left(\widehat{\theta}, V(\widehat\theta)\right)\).
\item
  For each sample, compute the survival function using Equation \ref{eqn:surv} below.
\item
  Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.
\end{enumerate}

This procedure is similar to parametric bootstrap (Efron and Tibshirani 1994, sec. 6.5), but it can be more accurately described as a form of approximate Bayesian computation. As such, the validity of the confidence bands relies on the Bernstein-von Mises theorem (Van der Vaart 2000, vol. 3, chap. 10).

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation}
In this model, the instantaneous failure rate is independent of \(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \emph{memoryless property} (Kalbfleisch and Prentice 2011).}

@@

Use of \(\log(t)\) yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice 2011):

@@

\[\frac{\lambda_j(t)^{\,\textrm dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
where \(N_j(t)\) is the counting process associated with the event of type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression with an offset term.\footnote{Specifically, it corresponds to the following parametrization for a multinomial random variable \(Y\): \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}

@@

Survival analysis is an important branch of applied statistics and epidemiology. Accordingly, there is already a vast ecosystem of R packages implementing different methodologies. In this section, we describe how the functionalities of \CRANpkg{casebase} compare to these packages.

At the time of writing, a cursory examination of CRAN's \ctv{Survival} Task View reveals that there are over 250 packages related to survival analysis (Allignol and Latouche 2019). For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and Cumulative Incidence (CI) estimation. By searching for appropriate keywords in the \texttt{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multistate models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.

Parametric survival models are implemented in several packages, each differing in the parametric distributions available: \CRANpkg{CFC} (2019), \CRANpkg{flexsurv} (2016), \CRANpkg{SmoothHazard} (2017), \CRANpkg{rstpm2} (2019), \CRANpkg{mets} (2014), and \CRANpkg{survival} (2015). For example, \pkg{SmoothHazard} is limited to Weibull distributions (2017), whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. \pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials. Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.

Several packages implement penalized estimation for the Cox model: \CRANpkg{glmnet} (2011), \CRANpkg{glmpath} (2018), \CRANpkg{penalized} (2010), \CRANpkg{riskRegression} (2020). Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{rstpm2} (2019), while \pkg{survival} (2015) has an implementation of ridge-penalized estimation. On the other hand, our package \pkg{casebase} provides penalized estimation of the hazard function. To our knowledge, \pkg{casebase} and \pkg{rstpm2} are the only packages to offer this functionality.

Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani \& Sharabiani (2019). The package \pkg{survival} provides functionality for competing risk analysis and multistate modelling. The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model (1999). On the other hand, the package \pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric survival functions. Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIs.

@@

\begin{table}

\caption{\label{tab:surv-pkgs}Comparison of various R packages for survival analysis. \textbf{Competing Risks}: whether an implementation for competing risks is present. \textbf{Allows Non PH}: includes models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: allows a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Risk estimates}: estimation of survival curve and cumulative incidence is available.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllllllll}
\toprule
\textbf{Package} & \textbf{Competing Risks} & \textbf{Allows Non PH} & \textbf{Penalized Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi Parametric} & \textbf{Interval/Left Censoring} & \textbf{Risk Estimates}\\
\midrule
casebase & X & X & X & X & X &  &  & X\\
CFC & X & X &  &  & X &  &  & X\\
cmprsk & X &  &  &  &  & X &  & X\\
crrp & X &  & X &  &  & X &  & \\
fastcox &  &  & X &  &  & X &  & \\
\addlinespace
flexrsurv &  & X &  & X & X &  &  & X\\
flexsurv & X & X &  & X & X &  &  & X\\
glmnet &  &  & X &  &  & X &  & X\\
glmpath &  &  & X &  &  & X &  & \\
mets & X &  &  & X &  & X &  & X\\
\addlinespace
penalized &  &  & X &  &  & X &  & \\
riskRegression & X &  & X &  &  & X &  & X\\
rstpm2 &  & X &  & X & X & X & X & X\\
SmoothHazard &  & X &  & X & X &  & X & \\
survival & X & X &  &  & X & X & X & X\\
\bottomrule
\end{tabular}}
\end{table}

@@

The functions in the \pkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function. We strove for compatibility with both \texttt{data.frame}s and \texttt{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.

@@

By separating the sampling and estimation functions, we allow the possibility of users implementing more complex sampling scheme (as described in Saarela (2016)), or more complex study designs (e.g.~time-varying exposure).
